---
title: "ETC3460 Assignment 1 Part 1"
format:
  pdf:
    fontsize: 11pt
    geometry: top=2cm, bottom=2cm, left=2.5cm, right=2.5cm
    number-sections: false
    include-before-body: 
      text: '\vspace{-1cm}'
---

Angus Ashby

## Question 1

a)  

```{r}
#| echo: false
MFST.mew <- 0.12
SBUX.mew <- 0.10

MSFT.sigma <- 0.3
SBUX.sigma <- 0.2
```

let cor =
0.0417$$0.0417 = \frac{Cov(MSFT,SBUX)}{0.3 \times 0.2}$$$$Cov(MSFT,SBUX) = 0.002502$$

```{r}
#| echo: false

# 1. Variable Assignments
cov.cov <- 0.002502

p1.mew <- SBUX.mew
p1.sigma <- SBUX.sigma
p2.mew <- (0.1 * MFST.mew) + (0.9 * SBUX.mew)  
p2.sigma <- sqrt((0.1^2 * MSFT.sigma^2) + (0.9^2 * SBUX.sigma^2) + 2*(0.1*0.9 * cov.cov))
p3.mew <- (0.2 * MFST.mew) + (0.8 * SBUX.mew)  
p3.sigma <- sqrt((0.2^2 * MSFT.sigma^2) + (0.8^2 * SBUX.sigma^2) + 2*(0.2*0.8 * cov.cov))
p4.mew <- (0.3 * MFST.mew) + (0.7 * SBUX.mew)  
p4.sigma <- sqrt((0.3^2 * MSFT.sigma^2) + (0.7^2 * SBUX.sigma^2) + 2*(0.3*0.7 * cov.cov))
p5.mew <- (0.4 * MFST.mew) + (0.6 * SBUX.mew)  
p5.sigma <- sqrt((0.4^2 * MSFT.sigma^2) + (0.6^2 * SBUX.sigma^2) + 2*(0.4*0.6 * cov.cov))
p6.mew <- (0.5 * MFST.mew) + (0.5 * SBUX.mew)  
p6.sigma <- sqrt((0.5^2 * MSFT.sigma^2) + (0.5^2 * SBUX.sigma^2) + 2*(0.5*0.5 * cov.cov))
p7.mew <- (0.6 * MFST.mew) + (0.4 * SBUX.mew)  
p7.sigma <- sqrt((0.6^2 * MSFT.sigma^2) + (0.4^2 * SBUX.sigma^2) + 2*(0.6*0.4 * cov.cov))
p8.mew <- (0.7 * MFST.mew) + (0.3 * SBUX.mew)  
p8.sigma <- sqrt((0.7^2 * MSFT.sigma^2) + (0.3^2 * SBUX.sigma^2) + 2*(0.7*0.3 * cov.cov))
p9.mew <- (0.8 * MFST.mew) + (0.2 * SBUX.mew)  
p9.sigma <- sqrt((0.8^2 * MSFT.sigma^2) + (0.2^2 * SBUX.sigma^2) + 2*(0.2*0.8 * cov.cov))
p10.mew <- (0.9 * MFST.mew) + (0.1 * SBUX.mew)  
p10.sigma <- sqrt((0.9^2 * MSFT.sigma^2) + (0.1^2 * SBUX.sigma^2) + 2*(0.9*0.1 * cov.cov))
p11.mew <- MFST.mew
p11.sigma <- MSFT.sigma

# 2. Data Aggregation
portfolio_results <- data.frame(
  Portfolio = 1:11,
  Expected_Return = c(p1.mew, p2.mew, p3.mew, p4.mew, p5.mew, p6.mew, p7.mew, p8.mew, p9.mew, p10.mew, p11.mew),
  S.Deviation = c(p1.sigma, p2.sigma, p3.sigma, p4.sigma, p5.sigma, p6.sigma, p7.sigma, p8.sigma, p9.sigma, p10.sigma, p11.sigma)
)

# 3. Table Rendering
knitr::kable(portfolio_results, caption = "Portfolio Expected Returns and Standard Deviation")
```

b)  The Global minimum Variance (GMV) portfolio defined as the lowest
    variance portfolio. Based purely on answers to Part a with 11
    portfolios to choose from, the GMV is given by portfolio 4 with 70%
    weight on SBUX and 30% weight on MSFT giving a variance of 0.0288

c)  

let cor =
-1$$-1 = \frac{Cov(MSFT,SBUX)}{0.3 \times 0.2}$$$$Cov(MSFT,SBUX) = -0.06$$

```{r}
#| echo: false

# 1. Variable Assignments (Execution occurs silently)
cov.cov <- -0.06

p1.mew <- SBUX.mew
p1.sigma <- SBUX.sigma
p2.mew <- (0.1 * MFST.mew) + (0.9 * SBUX.mew)  
p2.sigma <- sqrt((0.1^2 * MSFT.sigma^2) + (0.9^2 * SBUX.sigma^2) + 2*(0.1*0.9 * cov.cov))
p3.mew <- (0.2 * MFST.mew) + (0.8 * SBUX.mew)  
p3.sigma <- sqrt((0.2^2 * MSFT.sigma^2) + (0.8^2 * SBUX.sigma^2) + 2*(0.2*0.8 * cov.cov))
p4.mew <- (0.3 * MFST.mew) + (0.7 * SBUX.mew)  
p4.sigma <- sqrt((0.3^2 * MSFT.sigma^2) + (0.7^2 * SBUX.sigma^2) + 2*(0.3*0.7 * cov.cov))
p5.mew <- (0.4 * MFST.mew) + (0.6 * SBUX.mew)  
p5.sigma <- sqrt((0.4^2 * MSFT.sigma^2) + (0.6^2 * SBUX.sigma^2) + 2*(0.4*0.6 * cov.cov))
p6.mew <- (0.5 * MFST.mew) + (0.5 * SBUX.mew)  
p6.sigma <- sqrt((0.5^2 * MSFT.sigma^2) + (0.5^2 * SBUX.sigma^2) + 2*(0.5*0.5 * cov.cov))
p7.mew <- (0.6 * MFST.mew) + (0.4 * SBUX.mew)  
p7.sigma <- sqrt((0.6^2 * MSFT.sigma^2) + (0.4^2 * SBUX.sigma^2) + 2*(0.6*0.4 * cov.cov))
p8.mew <- (0.7 * MFST.mew) + (0.3 * SBUX.mew)  
p8.sigma <- sqrt((0.7^2 * MSFT.sigma^2) + (0.3^2 * SBUX.sigma^2) + 2*(0.7*0.3 * cov.cov))
p9.mew <- (0.8 * MFST.mew) + (0.2 * SBUX.mew)  
p9.sigma <- sqrt((0.8^2 * MSFT.sigma^2) + (0.2^2 * SBUX.sigma^2) + 2*(0.2*0.8 * cov.cov))
p10.mew <- (0.9 * MFST.mew) + (0.1 * SBUX.mew)  
p10.sigma <- sqrt((0.9^2 * MSFT.sigma^2) + (0.1^2 * SBUX.sigma^2) + 2*(0.9*0.1 * cov.cov))
p11.mew <- MFST.mew
p11.sigma <- MSFT.sigma

# 2. Data Aggregation
# Compiling the isolated variables into a structured data frame
portfolio_results <- data.frame(
  Portfolio = 1:11,
  Expected_Return = c(p1.mew, p2.mew, p3.mew, p4.mew, p5.mew, p6.mew, p7.mew, p8.mew, p9.mew, p10.mew, p11.mew),
  S.Deviation = c(p1.sigma, p2.sigma, p3.sigma, p4.sigma, p5.sigma, p6.sigma, p7.sigma, p8.sigma, p9.sigma, p10.sigma, p11.sigma)
)

# 3. Table Rendering
knitr::kable(portfolio_results, caption = "Portfolio Expected Returns and Standard Deviations")
```

d)  The Global minimum Variance (GMV) portfolio defined as the lowest
    variance portfolio. Based purely on answers to Part c with 11
    portfolios to choose from, the GMV is given by portfolio 5 with 60%
    weight on SBUX and 40% weight on MSFT giving a variance 6.9389 \*
    10\^-18

e)  

Note: $$W = X_a^2$$ ![Handwritten
Answer](C:/Users/angus/Downloads/pdf2png/Adobe%20Scan%2010%20Apr%202026/Adobe%20Scan%2010%20Apr%202026-1.png)
From the equation 2 we can calculate the weights of the GMV portfolio.

```{r}
#| include: false
MSFT.w <- (SBUX.sigma^2 - cov.cov)/(MSFT.sigma^2 + SBUX.sigma^2 - 2*cov.cov)
SBUX.w <- (1 - MSFT.w)
```

$$W_{MSFT} = \frac{\sigma_{SBUX}^2 - Cov}{\sigma_{MSFT}^2 + \sigma_{SBUX}^2 - 2(Cov)}$$$$W_{MSFT} = \frac{`r SBUX.sigma^2` - `r cov.cov`}{`r MSFT.sigma^2` + `r SBUX.sigma^2` - 2(`r cov.cov`)} = `r MSFT.w`$$$$W_{SBUX} = 1 - W_{MSFT} = 1 - `r MSFT.w` = `r SBUX.w`$$
The weights for the GMV portfolio given by this equation are 40% MSFT
and 60% SBUX, supporting the answer to part d.

## Question 2

a)  

```{r}
#| echo: false
#| message: false
#| warning: false

library("quantmod")
options("getSymbols.warning4.0"=FALSE)
getSymbols(Symbols=c("NVDA", 'WAT', 'JPM', 'L'), from="2020-12-01", to="2025-12-31",
           auto.assign=TRUE, warnings=FALSE)
          
NVDA <- NVDA$NVDA.Adjusted
WAT <- WAT$WAT.Adjusted
JPM <- JPM$JPM.Adjusted
L <- L$L.Adjusted

prices <- merge(NVDA, WAT, JPM, L)

prices.M <- to.monthly(prices, OHLC=FALSE)
colnames(prices.M) <- c("NVDA", "WAT", "JPM", "L")

prices.M.ret <- prices.M / lag(prices.M, 1) - 1
prices.M.ret <- na.omit(prices.M.ret)
equity.curve.M <- cumprod(1 + prices.M.ret)

# Plotting Execution
plot.zoo(equity.curve.M, 
         screens = 1,                 
         col = 1:4,                   
         xlab = "Date", 
         ylab = "Cumulative Return", 
         main = "Monthly Equity Curve")

legend("topleft", 
       legend = colnames(equity.curve.M), 
       col = 1:4,                     
       lty = 1,                       
       cex = 0.8)
```

b)  

```{r}
#| echo: false
#| results: asis
# Monthly Log Returns
NVDA.M.l.ret <- na.omit(diff(log(prices.M$NVDA)))
WAT.M.l.ret  <- na.omit(diff(log(prices.M$WAT)))
JPM.M.l.ret  <- na.omit(diff(log(prices.M$JPM)))
L.M.l.ret    <- na.omit(diff(log(prices.M$L)))
# Merge
log.returns.all <- merge(NVDA.M.l.ret, WAT.M.l.ret, JPM.M.l.ret, L.M.l.ret)
colnames(log.returns.all) <- c("NVDA", "WAT", "JPM", "L")
# 2. Define the risk-free rate
rf_annual <- 0.04
rf_monthly_log <- log(1 + rf_annual) / 12
# Compute Mean, Standard deviation and Sharpe Ratio
mean_returns <- apply(log.returns.all, 2, mean)
sd_returns <- apply(log.returns.all, 2, sd)
sharpe_ratio <- (mean_returns - rf_monthly_log) / sd_returns
stats_summary <- data.frame(
  Mean = mean_returns,
  StdDev = sd_returns,
  SharpeRatio = sharpe_ratio
)

knitr::kable(stats_summary, caption = "Summary Statistics")
```

c)  

```{r}
#| echo: false

# Compute covariance and correlation
cov.mat <- cov(log.returns.all)
cor.mat <- cor(log.returns.all)

# Render the correlation matrix natively as a formatted table
knitr::kable(cor.mat, digits = 4, caption = "4x4 Sample Correlation Matrix of Monthly Log Returns")
```

d)  Compute GMV weight
    $$\mathbf{w} = \frac{\Sigma^{-1} \mathbf{1}}{\mathbf{1}^T \Sigma^{-1} \mathbf{1}}$$

```{r}
#| echo: false

# 1. Computation
ones <- rep(1, ncol(cov.mat))
inv.cov.mat <- solve(cov.mat)
w.gmv.portfolio <- (inv.cov.mat %*% ones) / as.numeric(t(ones) %*% inv.cov.mat %*% ones)

# 2. Data Aggregation
# Extracting the numeric values and mapping them to their respective asset names
gmv_weights_df <- data.frame(
  Asset = c("NVDA", "WAT", "JPM", "L"),
  Weight = as.numeric(w.gmv.portfolio)
)

# 3. Table Rendering
knitr::kable(gmv_weights_df, digits = 4, caption = "Optimal Weights for the GMV Portfolio")
```

e)  

```{r}
#| echo: false
#| results: asis
# Define the risk-free rate
rf_annual <- 0.04
rf_monthly_log <- log(1 + rf_annual) / 12
# Compute Mean, Standard deviation and Sharpe Ratio
asset_means <- apply(log.returns.all, 2, mean)
port_mean_return <- as.numeric(t(w.gmv.portfolio) %*% asset_means)
port_variance <- as.numeric(t(w.gmv.portfolio) %*% cov.mat %*% w.gmv.portfolio)
port_sd <- sqrt(port_variance)

port_sharpe_ratio <- (port_mean_return - rf_monthly_log) / port_sd
stats_summary_port <- data.frame(
  Portfolio_Mean = port_mean_return,
  Portfolio_StdDev = port_sd,
  Portfolio_Sharpe = port_sharpe_ratio
)

knitr::kable(stats_summary_port, caption = "Portfolio Summary Statistics")
```

f)  the 5% Value-at-Risk (VaR) for a one-month investment is calculated
    using the sample mean ($\mu$) and standard deviation ($\sigma$) of
    the monthly log returns, assuming a normal distribution. The 5%
    quantile of returns is multiplied by the initial investment of
    \$50,000:

$$VaR_{5\%} = (\mu + \sigma \cdot Z_{0.05}) \times 50,000$$

```{r}
#| echo: false

initial_investment <- 50000

# NVDA
NVDA.M.l.ret <- na.omit(NVDA.M.l.ret)
NVDA.xbar <- mean(NVDA.M.l.ret)
NVDA.sd <- sd(NVDA.M.l.ret)
NVDA.q.5 <- NVDA.xbar + NVDA.sd * qnorm(0.05)  
VaR.5.NVDA <- NVDA.q.5 * initial_investment

# JPM
JPM.M.l.ret <- na.omit(JPM.M.l.ret)
JPM.xbar <- mean(JPM.M.l.ret)
JPM.sd <- sd(JPM.M.l.ret)
JPM.q.5 <- JPM.xbar + JPM.sd * qnorm(0.05)  
VaR.5.JPM <- JPM.q.5 * initial_investment

# WAT
WAT.M.l.ret <- na.omit(WAT.M.l.ret)
WAT.xbar <- mean(WAT.M.l.ret)
WAT.sd <- sd(WAT.M.l.ret)
WAT.q.5 <- WAT.xbar + WAT.sd * qnorm(0.05)  
VaR.5.WAT <- WAT.q.5 * initial_investment

# L
L.M.l.ret <- na.omit(L.M.l.ret)
L.xbar <- mean(L.M.l.ret)
L.sd <- sd(L.M.l.ret)
L.q.5 <- L.xbar + L.sd * qnorm(0.05)  
VaR.5.L <- L.q.5 * initial_investment

# GMV Portfolio
GMVp.xbar <- port_mean_return
GMVp.sd <- port_sd
GMVp.q.5 <- GMVp.xbar + GMVp.sd * qnorm(0.05)  
VaR.5.GMVp <- GMVp.q.5 * initial_investment

# Data Aggregation
var_results <- data.frame(
  Asset = c("Nvidia (NVDA)", "JPMorgan Chase (JPM)", "Waters Corp (WAT)", "Loews Corp (L)", "GMV Portfolio"),
  VaR_5_Percent = c(VaR.5.NVDA, VaR.5.JPM, VaR.5.WAT, VaR.5.L, VaR.5.GMVp)
)

# Table Rendering
knitr::kable(var_results, 
             digits = 2, 
             format.args = list(big.mark = ","), 
             caption = "5% Value-at-Risk (VaR) for a $50,000 One-Month Investment")
```

g)  

The Capital Asset Pricing Model (CAPM) regression is estimated using the
generic expression:

$$R_{i,t} - R_{f,t} = \alpha_i + \beta_i(R_{m,t} - R_{f,t}) + \epsilon_{i,t}$$

Where idiosyncratic risk is represented by the standard deviation of the
residuals, $\sigma(\epsilon_i)$. The Fama-French market factors and
asset returns are scaled as percentages

```{r}
#| echo: false
#| message: false
#| warning: false

library(frenchdata)
library(xts) 

# 1. Data Retrieval and Formatting
ff3 <- download_french_data("Fama/French 3 Factors")
ff3_monthly_raw <- ff3$subsets$data[[1]]
ff3_monthly_raw$date <- as.yearmon(as.character(ff3_monthly_raw$date), format = "%Y%m")

ff3_xts <- xts(
  ff3_monthly_raw[, c("Mkt-RF","SMB","HML","RF")],   
  order.by = as.Date(ff3_monthly_raw$date)                 
)
colnames(ff3_xts) <- c("mkt_excess","smb","hml","rf")
index(ff3_xts) <- as.yearmon(index(ff3_xts))  

# Aligning period and scaling returns
ff3.MRK <- ff3_xts["2021-01-01/2025-12-31"]
lr.stock <- log.returns.all * 100

# 2. CAPM Regressions
capm.NVDA <- lm(lr.stock$NVDA - ff3.MRK$rf ~ ff3.MRK$mkt_excess)
capm.WAT <- lm(lr.stock$WAT - ff3.MRK$rf ~ ff3.MRK$mkt_excess)
capm.JPM <- lm(lr.stock$JPM - ff3.MRK$rf ~ ff3.MRK$mkt_excess)
capm.L <- lm(lr.stock$L - ff3.MRK$rf ~ ff3.MRK$mkt_excess)

# GMV Portfolio Regression
w <- w.gmv.portfolio
gmv.ret.xts <- lr.stock[, "NVDA"]*w[1] + lr.stock[, "WAT"]*w[2] + lr.stock[, "JPM"]*w[3] + lr.stock[, "L"]*w[4] 
colnames(gmv.ret.xts) <- c("gmv_ret") 
capm.GMV <- lm(gmv.ret.xts - ff3.MRK$rf ~ ff3.MRK$mkt_excess)

# 3. Data Extraction and Aggregation
extract_metrics <- function(model, asset_name) {
  mod_summary <- summary(model)
  data.frame(
    Asset = asset_name,
    Alpha = mod_summary$coefficients[1, 1],
    SE_Alpha = mod_summary$coefficients[1, 2],
    Beta = mod_summary$coefficients[2, 1],
    SE_Beta = mod_summary$coefficients[2, 2],
    Idiosyncratic_Risk = mod_summary$sigma,
    R_Squared = mod_summary$r.squared
  )
}

capm_results_df <- rbind(
  extract_metrics(capm.NVDA, "NVDA"),
  extract_metrics(capm.WAT, "WAT"),
  extract_metrics(capm.JPM, "JPM"),
  extract_metrics(capm.L, "L"),
  extract_metrics(capm.GMV, "GMV Portfolio")
)

# 4. Table Rendering
knitr::kable(capm_results_df, 
             digits = 4, 
             caption = "CAPM Regression Output: Alpha, Beta, Risk, and Fit Metrics")
```

h)  **Historical Trajectories & Cross-Asset Correlation** The monthly
    equity curves shown in Part a plot the past return performance of
    each asset. The mathematical foundation for combining these assets
    is established in the correlation matrix (Part c). In a real-world
    context, the degree to which these four specific stocks move
    independently dictates our ability to reduce aggregate volatility
    without sacrificing expected returns. For instance, the correlation
    between NVDA and WAT is notably low (0.0939), providing significant
    diversification utility within the portfolio.

**Global Minimum Variance (GMV) and Downside Risk** Matrix optimization
(Part d) isolates the lowest-risk allocation possible for this
portfilio. The weightings are heavily skewed toward the least volatile
asset (L):

-   **NVDA:** 4.87%, **WAT:** 2.06%, **JPM:** 3.55%, **L:** 89.53%

To quantify downside risk, the 5% Value-at-Risk (VaR) translates
standard deviation into maximum expected 1-month dollar losses for a
\$50,000 investment at a 95% confidence interval (Part f):

| Asset                | 5% VaR (One-Month) |
|:---------------------|:-------------------|
| Nvidia (NVDA)        | -\$9,527.36        |
| Waters Corp (WAT)    | -\$7,370.32        |
| JPMorgan Chase (JPM) | -\$4,708.34        |
| Loews Corp (L)       | -\$3,260.28        |
| **GMV Portfolio**    | **-\$3,132.93**    |

The results demonstrate the practical effect of optimization: the GMV
portfolio restricts extreme downside dollar risk more effectively than
holding any single asset individually, even the defensive Loews Corp.

\*\* Asset Pricing and Systematic Risk Exposures \*\* The CAPM
regressions isolate fundamental pricing metrics for each equity,
defining how they relate to the broader market:

-   **Beta (**$\beta$): Measures systematic risk. **NVDA** exhibits the
    highest beta at **2.2919**, indicating aggressive sensitivity to
    market swings. Conversely, **L** is the most defensive with a
    $\beta$ of **0.5779**.\
-   **Idiosyncratic Risk & Fit (**$R^2$): **L** displays the lowest
    $R^2$ (**0.2783**). This low correlation to the market suggests its
    price action is driven predominantly by company-specific events
    (idiosyncratic risk) rather than macroeconomic trends.\
-   **Alpha (**$\alpha$): Evaluating the standard errors against the
    estimated alphas reveals that no asset or portfolio produced a
    statistically significant alpha at the 5% level (e.g., NVDA
    $t \approx 1.65$). This indicates that historical returns were
    structurally consistent with the systemic market risk taken, with no
    evidence of abnormal "beat-the-market" performance.

**Conclusion** The underlying data confirms that while individual
equities like NVDA provide high historical growth, they carry severe VaR
parameters and extreme systematic market exposure. Implementing the
weights of the GMV portfolio successfully counteracts these extremes,
anchoring the allocation in defensive assets to result in a highly
risk-adjusted profile that fundamentally minimizes capital drawdown.

## Question 3

a)  The Fama-French three factor (FF3F) regression model over the
    following subperiods is as follows:

```{r}
#| include: false

# 1. Data Retrieval and Preparation
ind5 <- download_french_data("5 Industry Portfolios")
id5.df <- ind5$subsets$data[[1]] 
id5.df$date <- as.yearmon(as.character(id5.df$date), format = "%Y%m")

id5.xts <- xts(
  id5.df[, c("Cnsmr","Manuf","HiTec","Hlth","Other")],   
  order.by = as.Date(id5.df$date)                  
)
index(id5.xts) <- as.yearmon(index(id5.xts))  

id5.manuf <- id5.xts$Manuf 
index(ff3_xts) <- as.yearmon(index(ff3_xts))  

# 2. Period 1: January 1997 to December 2011
id5.manuf.period1 <- id5.manuf["1997-01/2011-12"]
ff3.p1 <- ff3_xts["1997-01/2011-12"]

# FF3F Regression 1
ff3.fit1 <- lm(id5.manuf.period1 - ff3.p1$rf ~ ff3.p1$mkt_excess + ff3.p1$smb + ff3.p1$hml)

# 3. Period 2: January 2012 to December 2025
id5.manuf.period2 <- id5.manuf["2012-01/2025-12"]
ff3.p2 <- ff3_xts["2012-01/2025-12"]

# FF3F Regression 2
ff3.fit2 <- lm(id5.manuf.period2 - ff3.p2$rf ~ ff3.p2$mkt_excess + ff3.p2$smb + ff3.p2$hml)
```

January 1997 to December 2011

The fitted Fama-French three-factor regression for the first subperiod
is reported below:

```{r}
#| echo: false
# Extract coefficients and rename rows for clarity
coef_p1 <- summary(ff3.fit1)$coefficients
rownames(coef_p1) <- c("Intercept (Alpha)", "Mkt-RF (Beta)", "SMB", "HML")

knitr::kable(coef_p1, 
             digits = 4, 
             caption = "FF3F Regression Coefficients: Jan 1997 - Dec 2011")
```

January 2012 to December 2025

The fitted Fama-French three-factor regression for the second subperiod
is reported below:

```{r}
#| echo: false
# Extract coefficients and rename rows for clarity
coef_p2 <- summary(ff3.fit2)$coefficients
rownames(coef_p2) <- c("Intercept (Alpha)", "Mkt-RF (Beta)", "SMB", "HML")

knitr::kable(coef_p2, 
             digits = 4, 
             caption = "FF3F Regression Coefficients: Jan 2012 - Dec 2025")
```

b)  From the results of the regression, Period 1 defines
    $\beta_1 = 0.85056$, that is to say 1% increase in the market's
    excess return, the Manufacturing portfolio's excess return increases
    by 0.85056%. For period 2 this is 0.928025%, indicating during
    period 2, the manuf portfolio is more more sensitive to market
    movements in Period 2.

To test weather this portfolio tracks the market at a 5% significance
level through the hypothesis test,

$H_0: \beta_1 = 1$ $H_1: \beta_1 \neq 1$

Sub-Periods 1 and 2, denoted by $t_1$ and $t_2$ respectively;

$$t_1 = \frac{0.85056 - 1}{0.03863} = -3.868, \quad t_2 = \frac{0.928025 - 1}{0.030734} = -2.342$$

The degrees of freedom are 176 and 164 respectively so The Critical
value for $t_1$ is $\pm 1.9735$ and $t_2$ is $\pm 1.9745$

Therefore for both sub periods 1 and 2 since $|t| > t_{Critical}$ we
reject the null, and conclude at a 5% significance level we the beta IS
statistically different from 1, therefore the Manufacturing portfolio
does not perfectly track the market from Jan 1997 to Dec 2011, nor from
Jan 2012 to Dec 2025

c)  

```{r}
#| echo: false
# Restricted Regression Period 1&2:
capm_restricted.fit1 <- lm(id5.manuf.period1-ff3.p1$rf ~ ff3.p1$mkt_excess)
knitr::kable(summary(capm_restricted.fit1)$coefficients, digits = 4, caption = "CAPM Regression Coefficients: Jan 1997 - Dec 2011")

capm_restricted.fit2 <- lm(id5.manuf.period2-ff3.p2$rf ~ ff3.p2$mkt_excess)
knitr::kable(summary(capm_restricted.fit2)$coefficients, digits = 4, caption = "CAPM Regression Coefficients: Jan 2012 - Dec 2025")
```

**Period 1 Test:**

To test for the validity of the CAPM over the FF3F model using the
J-test at a 5% significance level:

$$H_0 : \beta_{SMB} = \beta_{HML} = 0$$
$$H_1 : \text{at least one is non-zero}$$

Under CAPM: $$2.659 = \sqrt{\frac{RSS_0}{178}}$$ $$RSS_0 = 1258.51$$

and under French Fama 3 factor: $$2.465 = \sqrt{\frac{RSS_1}{176}}$$
$$RSS_1 = 1069.4156$$

Hence J-Test: $$J = \frac{RSS_0 - RSS_1}{RSS_1 / (T-K-1)}$$
$$J = \frac{1258.51 - 1069.4156}{1069.4156 / 176}$$ $$J = 31.120$$

**Period 2 Test:**

To test for the validity of the CAPM over the FF3F model using the
J-test at a 5% significance level:

$$H_0 : \beta_{SMB} = \beta_{HML} = 0$$
$$H_1 : \text{at least one is non-zero}$$

Under CAPM: $$2.152 = \sqrt{\frac{RSS_0}{166}}$$ $$RSS_0 = 768.7633$$

and under French Fama 3 factor: $$1.579 = \sqrt{\frac{RSS_1}{164}}$$
$$RSS_1 = 408.8915$$

Hence J-Test: $$J = \frac{RSS_0 - RSS_1}{RSS_1 / (T-K-1)}$$
$$J = \frac{768.7633 - 408.8915}{408.8915 / 164}$$ $$J = 144.339$$

At a 5% level of significance, the critical value for a Chi-squared
distribution with 2 degrees of freedom is 5.991. The computed J-Test
statistics for period 1 is 31.120 and for Period 2, 144.339. Both are
greater than the Critical value therefore we reject the null hypothesis
($H_0: \beta_{SMB} = \beta_{HML} = 0$) in both sub periods. That is we
have sufficient evidence to suggest that the Fama-French 3-Factor does
not reduce to CAPM during either the 1997–2011 and 2011-2025 sub
periods.

d)  

The client’s claim is fundamentally flawed due to two primary
analytically errors: ignoring statistical significance and relying on a
miss specified model.

Whilst the Client correctly points out the CAPM intercept, or alpha for
the 1997-2011 is positive, (0.34895), the validity of this fails when
evaluated using a t-test to show if this value is statistically
significant compared to 0. At a 5% significance level, the p-value of
0.0808 \> 0.05, therefore we would fail to reject the null hypothesis
that alpha is 0, (H0: a=0). The observed deviation from zero is within
the expected range of random noise, meaning there is no evidence of true
excess returns over the market risk taken from the portfolio.

Additionally the client bases their buy signal from the alpha derived
from a purely CAPM model. However From the tests conducted earlier, for
the 1997-2011 sub period, we definitely rejected the null hypothesis
that the Fama-French 3-Factor (FF3F) model reduces to the CAPM. (see
Section 3 part C)

This proves CAPM regression calculated is is an inadequate formulation.
If we evaluate the portfolio's performance however using the FF3F model
for period 1, alpha = 0.27445, with a p-value of 0.141. Since this
p-value is again greater than 0.05 we cannot conclude the portfolio
generated statistically significant abnormal returns.

**Recommendation:** Do not buy based on the positive estimated
alpha-risk.

The decision to initiate a buy cannot be statistically support at a 5%
significance level from the regression models of CAPM or FF3F results,
however this does not indicate whether the trade is good or bad, as it
cannot be proved or disproved if buying will produce significant alpha
over excess market returns.

\newpage

## Appendix: R Code

```{r}
#| ref.label: !expr knitr::all_labels()
#| echo: true
#| eval: false
```
